---
title: "Facets"
output: pdf_document
date: "2023-07-12"
---
Faceting splits the data up into groups, according to a categorical variable, 
then plots each group in its own panel.
```{r}
# Load the libraries
library(ggplot2)
```
Facet layer basics:
```{r}
# Facet the plot in a grid, with each am value in its own row.
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point() +
  # Facet rows by am
  facet_grid(rows = vars(am))
```

```{r}
# Facet the plot in a grid, with each cyl value in its own column.
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point() +
  # Facet columns by cyl
  facet_grid(cols = vars(cyl))
```

```{r}
# Facet the plot in a grid, with each am value in its own row and each cyl 
# value in its own column.
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point() +
  # Facet rows by am and columns by cyl
  facet_grid(rows = vars(am),
  cols = vars(cyl))
```
Many variables:
```{r}
# Combine fcyl and fam into a single interaction variable, fcyl_fam
mtcars$fcyl_fam <- interaction(mtcars$fcyl, mtcars$fam)

# Color the points by fcyl_fam
ggplot(mtcars, aes(x = wt, y = mpg, colour = fcyl_fam)) +
  geom_point() +
  # Use a paired color palette
  scale_color_brewer(palette = "Paired")
```

```{r}
# Update the plot to map disp to size
ggplot(mtcars, aes(x = wt, y = mpg, color = fcyl_fam, size = disp)) +
  geom_point() +
  scale_color_brewer(palette = "Paired")
```

```{r}
# Update the plot
ggplot(mtcars, aes(x = wt, y = mpg, color = fcyl_fam, size = disp)) +
  geom_point() +
  scale_color_brewer(palette = "Paired") +
  # Grid facet on gear and vs
  facet_grid(rows = vars(gear), 
  cols = vars(vs))
```
Labeling facets:
```{r}
# Plot wt by mpg
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  # The default is label_value
  facet_grid(cols = vars(cyl))

# Plot wt by mpg
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  # Displaying both the values and the variables
  facet_grid(cols = vars(cyl), labeller = label_both)

# Plot wt by mpg
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  # Label context
  facet_grid(cols = vars(cyl), labeller = label_context)

# Plot wt by mpg
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  # Two variables
  facet_grid(cols = vars(vs, cyl), labeller = label_context)
```
Setting order:
```{r}
# Make factor, set proper labels explictly
mtcars$fam <- factor(mtcars$am, labels = c(`0` = "Automatic",
                                           `1` = "Manual"))

# Default order is alphabetical
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  facet_grid(cols = vars(fam))
```
Setting order part 2:
```{r}

# Manually set the label order
mtcars$fam <- factor(mtcars$am,
                     levels = c(1, 0),
                     labels = c("manual",
                      "automatic"))

# View again
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  facet_grid(cols = vars(fam))
```
Variable plotting spaces part 1: continuous variables
If the data ranges vary wildly between facets, it can be clearer if each facet 
has its own scale. This is achieved with the scales argument to facet_grid().

"fixed" (default): axes are shared between facets.
free: each facet has its own axes.
free_x: each facet has its own x-axis, but the y-axis is shared.
free_y: each facet has its own y-axis, but the x-axis is shared.
```{r}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() + 
  # Facet columns by cyl 
  facet_grid(cols = vars(cyl))

ggplot(mtcars, aes(wt, mpg)) +
  geom_point() + 
  # Update the faceting to free the x-axis scales
  facet_grid(cols = vars(cyl),
  scales = "free_x")

ggplot(mtcars, aes(wt, mpg)) +
  geom_point() + 
  # Swap cols for rows; free the y-axis scales
  facet_grid(rows = vars(cyl), scales = "free_y")
```
Variable plotting spaces part 2: categorical variables
```{r}
# Make rownames into a column as we will be using it for colour aes
mtcars <- tibble::rownames_to_column(mtcars, "car")

ggplot(mtcars, aes(x = mpg, y = car, color = fam)) +
  geom_point() +
  # Facet rows by gear
  facet_grid(rows = vars(gear))

ggplot(mtcars, aes(x = mpg, y = car, color = fam)) +
  geom_point() +
  # Free the y scales and space to remove blank lines
  facet_grid(rows = vars(gear),
  scales = "free_y",
  space = "free_y")
```







